---
title: "EC50_Lab1"
output: html_document
date: "2023-02-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Documents/R/EC50/Lab1")
library(tidyverse)
library(haven)
library(ggthemes)
library(statar) 
library(tidyquant)
nlsy97 <- read_dta("nlsy97.dta")
view(nlsy97)
```

```{r q1}
income_histogram <- nlsy97 |> 
                    ggplot() +
                    geom_histogram(aes(kid_income)) +
                    labs(x = "Child income, averaged between 2013 and 2015") +
                    scale_x_continuous(labels = scales::dollar_format())
income_histogram
  
```

```{r q2}

mean_kidincome <- mean(nlsy97$kid_income, na.rm = TRUE)
mean_kidincome

```

```{r q3}

nlsy97 <- nlsy97 |> 
  mutate(below_mean = if_else(nlsy97$kid_income < mean_kidincome, 1, 0))

mean(nlsy97$below_mean)




```

The sample mean of `below_mean` is .596, suggesing that just under 60% of the observations have a value of `kid_income` below its mean. This is because the dataset is skewed with a number of outlier values for `kid_income`, as seen in the histogram with the high number of right-side values of over $300,000. 











## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
