---
title: "EC50_Lab1"
output:
  pdf_document: default
  html_document: default
date: "2023-02-04"
---

```{r setup, include=FALSE}
#Installing required packages and reading in the data 
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Documents/R/EC50/Lab1")
library(tidyverse)
library(haven)
library(ggthemes)
library(statar) 
library(tidyquant)
nlsy97 <- read_dta("nlsy97.dta")
view(nlsy97)
```

```{r q1}
#Creating histogram of kid_income
income_histogram <- nlsy97 |> 
                    ggplot() +
                    geom_histogram(aes(kid_income)) +
                    labs(x = "Child income, averaged between 2013 and 2015") +
                    scale_x_continuous(labels = scales::dollar_format())
income_histogram
  
```


```{r q2}
#Calculating mean of kid_income 
mean_kidincome <- mean(nlsy97$kid_income, na.rm = TRUE)
mean_kidincome

```
The sample mean of `kid_income` is $70499.94. 

```{r q3}

#Creating indicator variable for values of kid_income below the mean and calculating the proportion of observations below the mean of kid_income 
nlsy97 <- nlsy97 |> 
  mutate(below_mean = if_else(nlsy97$kid_income < mean_kidincome, 1, 0))

mean(nlsy97$below_mean)

```

The sample mean of `below_mean` is .596, suggesing that just under 60% of the observations have a value of `kid_income` below its mean. This is because the dataset is skewed with a number of outlier values for `kid_income`, as seen in the histogram with the high number of right-side values of over $300,000. 

```{r q4}
#Calculating the median of kid_income
median_kidincome <- median(nlsy97$kid_income, na.rm = TRUE)
median_kidincome

```


The sample median of `kid_income` is $58750. 

```{r q5}
#Calculating the standard deviation of kid_income
sd_kidincome <- sd(nlsy97$kid_income, na.rm = TRUE)
sd_kidincome

```

The sample standard deviation of `kid_income` is $59552.02. 

```{r q6}
#Creating indicator variables for values of kid_income within 1 and 2 SDs of the mean and calculating the proportion of observations that fall within 1 and 2 SDs of the mean of kid_income 
upper_bound1sd <- mean_kidincome + sd_kidincome
upper_bound1sd

lower_bound1sd <- mean_kidincome - sd_kidincome
lower_bound1sd

upper_bound2sd <- mean_kidincome + (sd_kidincome * 2)
upper_bound2sd

lower_bound2sd <- mean_kidincome - (sd_kidincome * 2)
lower_bound2sd

nlsy97 <- nlsy97 |> 
          mutate(within1sd = ifelse(kid_income <= upper_bound1sd & kid_income >= lower_bound1sd, 1, 0),
                 within2sd = ifelse(kid_income <= upper_bound2sd & kid_income >= lower_bound2sd, 1, 0))
nlsy97

mean(nlsy97$within1sd)
mean(nlsy97$within2sd)

```

Roughly 79% of observations are within one SD of the mean of `kid_income` and almost 95% of observations are within two SDs of the mean of `kid_income`. This sample is not quite normally distributed. 

```{r q7}
#Creating ranked percentiles for all observations of kid_income 
nlsy97 <- nlsy97 |> 
          mutate(ranked_kidincome = rank(kid_income))

max_rank <- max(nlsy97$ranked_kidincome)

nlsy97$kid_inc_rank <-  (nlsy97$ranked_kidincome/max_rank)*100

percentile_rank <- function(variable){
  r <- ifelse(is.na(variable), NA, rank(variable, ties.method = "average"))
  100*r/max(r, na.rm = TRUE)
}

nlsy97$kid_inc_rank <- with(nlsy97, percentile_rank(kid_income))
view(nlsy97)

#Displaying the top 10 and bottom 10 observations ranked by kid_inc_rank
toprank <- nlsy97 |> 
  arrange(desc(kid_inc_rank)) |> 
  slice(1:10)

bottomrank <- nlsy97 |> 
  arrange(kid_inc_rank) |> 
  slice(1:10)


```


```{r q8}
#Creating a histogram of kid_inc_rank, showcasing an approximately uniform distribution
rank_histo <- nlsy97 |> 
  ggplot() +
  geom_histogram(aes(x = kid_inc_rank, y = after_stat(density))) +
  labs(title = "Ranked child income",
       subtitle = "Averaged across 2013 and 2015",
       x = "Rank")

rank_histo

#Calculating the mean and median of kid_inc_rank
mean(nlsy97$kid_inc_rank)
median(nlsy97$kid_inc_rank)


```


As shown above, the sample mean and median for `kid_inc_rank` are approximately equal, 50.08 and 50.11. 

```{r q9}

#examining the relationship between kid income and parents' income

linearscatter <- nlsy97 |> 
  ggplot(aes(x = parent_inc, y = kid_income)) +
  stat_binmean(n = 20, geom = "point") 
linearscatter 

ggsave("linearscatter.png")

#examining the relationship between kid income and years of education 

nonlinearscatter <- nlsy97 |> 
  ggplot(aes(x = child_education, y = kid_income)) +
  stat_binmean(n = 20, geom = "point") 
nonlinearscatter 
ggsave("nonlinearscatter.png")

```

```{r q10}
#ensuring reproducibility
set.seed(21519588)

#creating treatment and control groups
nlsy97 <- nlsy97 |> 
  mutate(random_number = runif(length(parent_inc)),
         treatment_group = ifelse(random_number >= 0.5, 1, 0))

#Report total number of observations in treatment group
sum(nlsy97$treatment_group)
#Report total number of observations in control group
sum(1-nlsy97$treatment_group)

#Reporting summary statistics for all variables by treatment group
options(dplyr.width = Inf)
nlsy97 |> 
  group_by(treatment_group) |> 
  summarise_all("mean")
nlsy97 |> 
  group_by(treatment_group) |> 
  summarise_all("sd")

```


Random assignment reduces the risk of bias influencing the results of any experiment. Subconscious or conscious bias when allocating groups through human judgment could lead to groups that are not approximately equal, which can lead to confounding factors that contributes to the fundamental problem of causal inference. Bias could lead to less than verifiable or reproducible results that claim to show a causal link or effect between two or more variables. Reducing bias strengthens the internal validity of any experiment's results and makes for a more robust experiment. 





























